Fused regression for multi-source gene regulatory network inference

### 2016-08-04 paper_revisions.tex
hi

We develop two algorithms for solving efficiently multi-output least-squares regression problems with pairwise L2 fusion penalties on entries of the coefficient matrix, and discuss conditions under which each is suitable. We also introduce - in the form of adaptive fusion - the idea of a saturating penalty function on fusion constraints, and estimate the solution to the resulting optimization problem through iterative application of the fused L2 algorithm. We start with a discussion of the fused L2 and adaptive fusion algorithms, and describe their performance on synthetic datasets intended to represent related gene regulatory networks. We then demonstrate the applications of the fused L2 network inference algorithm on biological data, first demonstrating gains in cross-species network inference using the bacteria species \texit{Bacillus subtilis} and the distantly related \texti{Bacillus anthracis}, then moving on to multi-platform network inference using \texit{B. subtilis}, and finally intra-species fused network inference using priors based on operons in \textit{B. subtilis}. We then discuss adaptive fusion and show the technique's ability to identify incorrect orthology information which has been introduced to a biological dataset, suggesting the technique's applicability to discovering neo- and sub-functionalizations.

@@

And for steady state conditions
\begin{equation}
\alpha_{i}x_{i}(t_k) = \sum \beta_{i,j}x_{j}(t_k)
\end{equation}
Both kinds of conditions can be included in the set of equations to be solved. We can summarize these equations in matrix form as
\begin{equation}
Y = X \beta 
\end{equation}
where $Y$ is the gene expression matrix, $X$ is the TF expression matrix, and $\beta$ is the regulatory weights we are interested in learning.
We are interested in learning $\beta$, the matrix representation of the gene regulatory network, where the weight in a given position represents the regulatory weight of a TF on a gene. 
Positive weights represent activation, negative weights represent repression, and 0 weights represent the absence of an interaction. The matrix $\beta$ can be solved using linear regression. 
Because there are typically far fewer conditions than possible regressors (TFs), we introduce a ridge regularization constraint with weight $\lambda_R$ and solve
\begin{equation}
\argmin_\beta\vert \vert X\beta - Y_2 \vert \vert ^2 + \lambda_R \vert \vert \beta \vert \vert ^2
\end{equation}
This is similar to the formulation used in the \textit{Inferelator} algorithm, except that we use an $L2$ penalty in place of the more standard $L1$ penalty. Although $L1$ penalties have the theoretical appeal of sparsity - which matches our assumption of sparsity in biological networks - we are primarily interested here in ranking interactions, rather than recovering a specific network. Moreover, $L2$ penalized regression followed by thresholding can be competitive with $L1$ penalized regression for support recovery \cite{} [LOOK UP THIS REF]. Although our approach can be easily modified to use an $L1$ penalty, we did not observe performance gains when doing so, and the version presented here is in some ways conceptually simpler (with the combined penalty function being equivalent to a multivariate Gaussian prior).

@@

Although fused regression allows more accurate identification of pairs of non-conserved interactions, the weights obtained for these interactions will be biased towards one another. We have described a method -- adaptive fusion -- that attempts to address this bias by simultaneously learning the networks and constraint weights. This method is based on minimizing a saturating penalty function on fusion constraints, similar to a class of penalties that have been developed to minimize bias in regularized regression \cite{fan2001variable,zhang2010nearly}. 
The result of adaptive fusion is both a network and a new set of fusion constraints, the weights of which can be interpreted as describing conservation structure across the networks.For the multiple species case, relaxation of fusion constraints represents orthologs which do not share similar interactions presumably due to evolution of regulatory circuitry \cite{kellis_proof_2004}. %not sure what radid is
When jointly learning networks describing processes in different cell lines, this may identify interesting context-specific behavior. Genes may be fused together on the basis of similar binding sites or chromatin features, and the relaxing of the fusion penalty indicates divergence of gene function. [ZACK - CONSIDER CUTTING, SPECULATIVE]

@@

\begin{figure}
\begin{center}
  \includegraphics[scale=0.30]{fig1.pdf}
  \caption{\label{fusion} \textbf{A.} Schematic representation of the the generation of fusion constraints from orthology mappings. Dashed arrows indicate potential regulatory interactions, while solid arrows denote orthology. We introduce fusion constraints for pairs of interactions for which both the regulator and regulated gene are orthologs of one another.   In this example, we would introduce a constraint between the $(A, B)$ and $(A', B')$ interactions and the $(A, C)$ and $(A', B')$ interactions. \textbf{B.} 
  In order to demonstrate the utility of fused network inference in combining data, we generate two networks with 10 TFs and 200 genes (75\% sparsity). Mean squared error of the inferred vs. true coefficient matrices for network 1 are plotted as a function of the number of conditions generated for species 1 (x-axis) and the number of conditions generated for species 2 (y-axis). 
  As expected, increasing the number of samples available for the species of interest improves network inference performance. 
  However, because we are fusing to data from a related species, similar gains are observed when increasing the amount of data available in this second species. 
  \textbf{C-F} Show the varying effects of fusion on simulated networks with different levels of conservation. We generate a series of networks with 20 TFs by 200 genes in two species, (50\% sparsity) while varying the fraction of gene orthologies in the simulated networks (Amount fused, shown above each figure). For each network, we evaluated AUPR on one of the species for: all interactions (blue line), interactions with fusion constraints (green line), and interactions without fusion constraints (orange line). At every level of conservation, constrained interactions show the largest benefits of fusion, with the magnitude of the benefit growing with fraction of orthologous genes. 
  When the networks are highly conserved, however, even interactions that are not directly constrained through fusion are recovered more accurately as $\lambda_S$ increases.}
\end{center}
\end{figure}
\pagebreak

@@

\begin{figure}
\begin{center}
  \includegraphics[scale=0.30]{fig5_2.pdf}
  \caption{\label{xspecies-real} \textbf{A} We first optimized $\lambda_R$, then performed a 20-fold cross-validation procedure in which subsets of the \textit{B. subtilis} network were fit either independently or fused to the \textit{B. anthracis network}. Each \textit{B. subtilis} network was then evaluated on a gold standard of known interactions, to simulate results of network inference with a small amount of available data in the species of interest (\textit{B. subtilis}) but more data available in a related species (\textit{B. anthracis}). Fusion with $\lambda_S=1.0$ improved performance, yielding a mean AUPR across the learned networks of 0.0388 vs. 0.0298 fit independently of  \textit{B. anthracis}. Plotted is the resulting precision-recall curve, with significant performan gains at low values of recall. In \textbf{B.,C.} we tested adaptive fusion using the same procedure, adding an additional 561 randomly selected additional orthologs. We set the $a$ parameter of adaptive fusion equal to $60t$ percentile of the differences in fused interaction weights when networks were fit with $\lambda_S=0$, so that approximately $40\%$ of constraints were relaxed. This reflected our belief that many of the fusion constraints arising from known orthologies did not necessarily reflect functional similarity, and could be unfused. \textbf{B.} Shows the fraction of fusion constraints arising from known orthologies and false random orthologies which were relaxed by adaptive fusion; a larger fraction of constraints known to not reflect functional conservation were relaxed. \textbf{C.} After fitting the networks independently, we plot the distribution of the differences in weight of interactions fused by known false fusion constraints and fusion constraints arising from known orthology.}
\end{center}
\end{figure}
\pagebreak

>>>>

We develop two algorithms for solving efficiently multi-output least-squares regression problems with pairwise L2 fusion penalties on entries of the coefficient matrix, and discuss conditions under which each is suitable. We also introduce - in the form of adaptive fusion - the idea of a saturating penalty function on fusion constraints, and estimate the solution to the resulting optimization problem through iterative application of the fused L2 algorithm. We start with a discussion of the fused L2 and adaptive fusion algorithms, and describe their performance on synthetic datasets intended to represent related gene regulatory networks. We then demonstrate the applications of the fused L2 network inference algorithm on biological data, first demonstrating gains in cross-species network inference using the bacteria species \texit{Bacillus subtilis} and the distantly related \textit{Bacillus anthracis}, then moving on to multi-platform network inference using \texit{B. subtilis}, and finally intra-species fused network inference using priors based on operons in \textit{B. subtilis}. We then discuss adaptive fusion and show the technique's ability to identify incorrect orthology information which has been introduced to a biological dataset, suggesting the technique's applicability to discovering neo- and sub-functionalizations.

@@

And for steady state conditions
\begin{equation}
\alpha_{i}x_{i}(t_k) = \sum \beta_{i,j}x_{j}(t_k)
\end{equation}
Both kinds of conditions can be included in the set of equations to be solved. We can summarize these equations in matrix form as
\begin{equation}
Y = X \beta 
\end{equation}
where $Y$ is the gene expression matrix, $X$ is the TF expression matrix, and $\beta$ is the regulatory weights we are interested in learning.
We are interested in learning $\beta$, the matrix representation of the gene regulatory network, where the weight in a given position represents the regulatory weight of a TF on a gene. 
Positive weights represent activation, negative weights represent repression, and 0 weights represent the absence of an interaction. The matrix $\beta$ can be solved using linear regression. 
Because there are typically far fewer conditions than possible regressors (TFs), we introduce a ridge regularization constraint with weight $\lambda_R$ and solve
\begin{equation}
\argmin_\beta\vert \vert X\beta - Y_2 \vert \vert ^2 + \lambda_R \vert \vert \beta \vert \vert ^2
\end{equation}
This is similar to the formulation used in the \textit{Inferelator} algorithm, except that we use an $L2$ penalty in place of the more standard $L1$ penalty. Although $L1$ penalties have the theoretical appeal of sparsity - which matches our assumption of sparsity in biological networks - we are primarily interested here in ranking interactions, rather than recovering a specific network. Moreover, $L2$ penalized regression followed by thresholding can be competitive with $L1$ penalized regression for support recovery \cite{shao_estimation_2012}. Although our approach can be easily modified to use an $L1$ penalty, we did not observe performance gains when doing so, and the version presented here is in some ways conceptually simpler (with the combined penalty function being equivalent to a multivariate Gaussian prior).

@@

Although fused regression allows more accurate identification of pairs of non-conserved interactions, the weights obtained for these interactions will be biased towards one another. We have described a method -- adaptive fusion -- that attempts to address this bias by simultaneously learning the networks and constraint weights. This method is based on minimizing a saturating penalty function on fusion constraints, similar to a class of penalties that have been developed to minimize bias in regularized regression \cite{fan2001variable,zhang2010nearly}. 
The result of adaptive fusion is both a network and a new set of fusion constraints, the weights of which can be interpreted as describing conservation structure across the networks.For the multiple species case, relaxation of fusion constraints represents orthologs which do not share similar interactions presumably due to evolution of regulatory circuitry \cite{kellis_proof_2004}. %not sure what radid is
When jointly learning networks describing processes in different cell lines, this may identify interesting context-specific behavior. Genes may be fused together on the basis of similar binding sites or chromatin features, and the relaxing of the fusion penalty indicates divergence of gene function.

@@

\begin{figure}
\begin{center}
  \includegraphics[scale=0.30]{fig2_revisions.pdf}
  \caption{\label{fusion} \textbf{A.} Schematic representation of the the generation of fusion constraints from orthology mappings. Dashed arrows indicate potential regulatory interactions, while solid arrows denote orthology. We introduce fusion constraints for pairs of interactions for which both the regulator and regulated gene are orthologs of one another.   In this example, we would introduce a constraint between the $(A, B)$ and $(A', B')$ interactions and the $(A, C)$ and $(A', B')$ interactions. \textbf{B.} 
  In order to demonstrate the utility of fused network inference in combining data, we generate two networks with 10 TFs and 200 genes, 75\% of which are orthologous (75\% sparsity). Mean squared error of the inferred vs. true coefficient matrices for network 1 are plotted as a function of the number of conditions generated for species 1 (x-axis) and the number of conditions generated for species 2 (y-axis). 
  As expected, increasing the number of samples available for the species of interest improves network inference performance. 
  However, because we are fusing to data from a related species, similar gains are observed when increasing the amount of data available in this second species. 
  \textbf{C-F} Show the varying effects of fusion on simulated networks with different levels of conservation. We generate a series of networks with 20 TFs by 200 genes in two species, (50\% sparsity) while varying the fraction of gene orthologies in the simulated networks (Amount fused, shown above each figure). For each network, we evaluated AUPR on one of the species for: all interactions (blue line), interactions with fusion constraints (green line), and interactions without fusion constraints (orange line). At every level of conservation, constrained interactions show the largest benefits of fusion, with the magnitude of the benefit growing with fraction of orthologous genes. 
  When the networks are highly conserved, however, even interactions that are not directly constrained through fusion are recovered more accurately as $\lambda_S$ increases.}
\end{center}
\end{figure}
\pagebreak

@@

\begin{figure}
\begin{center}
  \includegraphics[scale=0.30]{fig5_3.pdf}
  \caption{\label{xspecies-real} \textbf{A} We first optimized $\lambda_R$, then performed a 20-fold cross-validation procedure in which subsets of the \textit{B. subtilis} network were fit either independently or fused to the \textit{B. anthracis network}. Each \textit{B. subtilis} network was then evaluated on a gold standard of known interactions, to simulate results of network inference with a small amount of available data in the species of interest (\textit{B. subtilis}) but more data available in a related species (\textit{B. anthracis}). Fusion with $\lambda_S=1.0$ improved performance, yielding a mean AUPR across the learned networks of 0.0388 vs. 0.0298 fit independently of  \textit{B. anthracis}. Plotted is the resulting precision-recall curve, with significant performance gains at low values of recall. We show 95\% bootstrap confidence interval. In \textbf{B.,C.} we tested adaptive fusion using the same procedure, adding an additional 561 randomly selected additional orthologs. We set the $a$ parameter of adaptive fusion equal to $60t$ percentile of the differences in fused interaction weights when networks were fit with $\lambda_S=0$, so that approximately $40\%$ of constraints were relaxed. This reflected our belief that many of the fusion constraints arising from known orthologies did not necessarily reflect functional similarity, and could be unfused. \textbf{B.} Shows the fraction of fusion constraints arising from known orthologies and false random orthologies which were relaxed by adaptive fusion; a larger fraction of constraints known to not reflect functional conservation were relaxed. \textbf{C.} After fitting the networks independently, we plot the distribution of the differences in weight of interactions fused by known false fusion constraints and fusion constraints arising from known orthology.}
\end{center}
\end{figure}
\pagebreak
